Data Analysis Methods for the Detection of the Epoch of Reionization

### 2017-07-17 PAPER.tex
fleshed out thermal noise parameter descriptions

Beginning with compressed data products which have been cleaned of radio frequency interference (RFI) at the $6\sigma$ level and then down-sampled (to $42.9$ s integrations, $203$ frequency channels), we then employ the technique of redundant calibration as a means of calibrating for individual antenna gains without needing to use information about the sky. Because PAPER is laid out on a grid, antenna calibration parameters can be found through the use redundancy. Namely, baselines of the same length and orientation should measure the same sky signal, but in practice there are differences due to instrumental effects caused by antennas, cables, and receivers. Using the basis of redundancy, however, allows us to set up a system of equations to solve for a complex gain per antenna that brings all baseline of a particular type into alignment. We perform these computations using the package OMNICAL \cc{citation needed}.

@@

Notice that there are two quantities of $\textbf{w}\textbf{x}$ in the expression for $\hat{\textbf{q}}$ in Equation \ref{eq:qhat}, representing two copies of weighted data. We perform all possible cross-multiplications of this quantity, except between the same datasets (`even' with `even', for example) and same groups (`baseline group 1' with `baseline group 1', for example). We avoid these computations to prevent the introduction of biases from shared auto power.

As mentioned previously, we have a choice for our normalization matrix $\textbf{M}$. For our analysis, we choose to compute it by taking the Cholesky decomposition of $\textbf{F}$. Namely, we decompose the Fisher matrix such that $\textbf{F} = \textbf{L}\textbf{L}^{\dagger}$, where $\textbf{L}$ is a lower triangular matrix. Next, we construct $\textbf{M} = \textbf{D}\textbf{L}^{-1}$, where \textbf{D} is a diagonal matrix. Hence, our window function $\textbf{W} = \textbf{MF}$ becomes $\textbf{W} = \textbf{D}\textbf{L}^{\dagger}$ and is an upper triangular matrix. This window function was constructed in a way to prevent the leakage of foreground power from low $k$ to high $k$ modes. Specifically, we order the elements in $\textbf{F}$ in such a way so that power can leak from high $k$ modes to low $k$ modes, but not vice versa. Since most foreground power shows up at low $k$'s, this method ensures a window function that retains clean, noise-dominated measurements.

@@

One way to estimate errors of a $21$ cm power spectrum estimate is to compute theoretical errors based on instrumental temperature and observational parameters. A theoretical noise limit provides a useful cross-check for bootstrapped errors and is a helpful predictor of the maximum sensitivity that can achieved by a particular observation and analysis. Here we walk through a detailed computation of a theoretical noise estimation as applied to PAPER-64 observations and highlight major changes from \citet{ali_et_al2015}. We first give a general formula, and then go into detail about each factor.

@@

\begin{itemize}
\item $X^{2}Y$: Conversion factors from observing coordinates (angles on the sky) to cosmological coordinates (co-moving distances). For $z=8.4$, $X^{2}Y = 5 \times 10^{11} [h^{-3} Mpc^{3} / str \, GHz]$.
\item $\Omega_{eff}$: The primary beam angular size. The effective beam area changes with the application of a fringe-rate filter, since parts of the beam are up-weighted and down-weighted. Using numbers from Table 1 in \citet{parsons_et_al2016}, $\Omega_{eff} = 0.74^{2}/0.24$ for an optimal fringe-rate filter. 
\item $T_{sys}$: The system temperature is set by Equation \ref{eq:sys}, for a receiver temperature of $200$ K and center frequency of $0.15$ GHz. 
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from only using the real part of power spectrum estimates when plotting and quoting upper limits.
\item $N_{lsts}$: The number of LST hours that go into power spectrum estimation. The sensitivity scales as the square root because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by... For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. 
\item $N_{bls}$:
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}

>>>>

Beginning with compressed data products which have been cleaned of radio frequency interference (RFI) at the $6\sigma$ level and then down-sampled (to $42.9$ s integrations, $203$ frequency channels), we then employ the technique of redundant calibration as a means of calibrating for individual antenna gains without needing to use information about the sky. Because PAPER is laid out on a grid, antenna calibration parameters can be found through the use redundancy. Namely, baselines of the same length and orientation should measure the same sky signal, but in practice there are differences due to instrumental effects caused by antennas, cables, and receivers. Using the basis of redundancy, however, allows us to set up a system of equations to solve for a complex gain per antenna that brings all baseline of a particular type into alignment. We perform these computations using the package OMNICAL (\citealt{zheng_et_al2014}).

@@

Notice that there are two quantities of $\textbf{w}\textbf{x}$ in the expression for $\hat{\textbf{q}}$ in Equation \ref{eq:qhat}, representing two copies of weighted data. We perform all possible cross-multiplications of this quantity, except between the same datasets (`even' with `even', for example) and same groups (`baseline group 1' with `baseline group 1', for example). We avoid these computations to prevent the introduction of noise biases from shared auto power.

As mentioned previously, we have a choice for our normalization matrix $\textbf{M}$. For our analysis, we choose to compute it by taking the Cholesky decomposition of $\textbf{F}$. Namely, we decompose the Fisher matrix such that $\textbf{F} = \textbf{L}\textbf{L}^{\dagger}$, where $\textbf{L}$ is a lower triangular matrix. Next, we construct $\textbf{M} = \textbf{D}\textbf{L}^{-1}$, where \textbf{D} is a diagonal matrix. Hence, our window function $\textbf{W} = \textbf{MF}$ becomes $\textbf{W} = \textbf{D}\textbf{L}^{\dagger}$ and is an upper triangular matrix. This window function was constructed in a way to prevent the leakage of foreground power from low $k$ to high $k$ modes. Specifically, we order the elements in $\textbf{F}$ in such a way so that power can leak from high $k$ modes to low $k$ modes, but not vice versa. Since most foreground power shows up at low $k$'s, this method ensures a window function that retains clean, noise-dominated measurements while minimizing foreground bias.

@@

One way to estimate errors on a $21$ cm power spectrum estimate is to compute theoretical errors based on the instrumental temperature and observational parameters. A theoretical noise limit provides a useful cross-check for bootstrapped errors and is a helpful predictor of the maximum sensitivity that can be achieved by a particular observation and analysis. Here we walk through a detailed computation of a theoretical noise estimation as applied to PAPER-64 observations and highlight major changes from \citet{ali_et_al2015}. We first give a general formula, and then go into detail about each factor.

@@

\begin{itemize}
\item $X^{2}Y$: Conversion factors from observing coordinates (angles on the sky) to cosmological coordinates (co-moving distances). For $z=8.4$, $X^{2}Y = 5 \times 10^{11} h^{-3} Mpc^{3} str^{-1} GHz^{-1}$.
\item $\Omega_{eff}$: The primary beam angular size. The effective beam area changes with the application of a fringe-rate filter, since parts of the beam are up-weighted and down-weighted. Using numbers from Table 1 in \citet{parsons_et_al2016}, $\Omega_{eff} = 0.74^{2}/0.24$ for an optimal fringe-rate filter. 
\item $T_{sys}$: The system temperature is set by Equation \ref{eq:sys}, for a receiver temperature of $200$ K and center frequency of $0.15$ GHz. \cc{does this differ from prior analysis?}
\item $\sqrt{2}$: This factor in the denominator of the sensitivity equation comes from only using the real part of power spectrum estimates when plotting and quoting upper limits (imaginary component is discarded).
\item $N_{lsts}$: The number of LST hours that go into a power spectrum estimation. The sensitivity scales as the square root because we integrate incoherently over time. For PAPER-64, $N_{lsts} = 8$ hours.
\item $N_{seps}$: The number of baseline separation types averaged incoherently in a final power spectrum estimate. For the analysis in this paper, we only use one type of baseline, hence $N_{seps}=1$.
\item $t_{int}$: The integration time of the data. It is crucial to adapt this number if filtering is applied along the time axis (i.e. a fringe-rate filter). We compute the effective integration time of our fringe-rate filtered data by scaling our original integration time using the following:
\begin{equation}
t_{frf} = t_{int} \frac{\int1 \, df}{\int w^{2}(f) \,df},
\end{equation}
where $t_{int}=43$ seconds, $t_{frf}$ is our fringe-rate filtered integration time, $w$ is our fringe-rate profile, and the integral is taken over all fringe-rates. In other words, we calculate our effective integration time by computing how our beam is altered in fringe-rate space. For PAPER-64, this number is $t_{int} = 3857$ s. 
\item $N_{days}$: The total number of days of data analyzed. In \citet{ali_et_al2015}, this number was set to $135$. However, because we divide our data in half (to form `even' and `odd' datasets), this number should be reduced by a factor of $2$. Additionally, because our $LST$ coverage is not $100\%$ complete (it doesn't overlap for every single day), we compute a realistic value of $N_{days}$ as the RMS of all the daily counts in one dataset (`even', for example). For PAPER-64, our revised estimate of $N_{days}$ is $\sim34$ days.
\item $N_{bls}$: The number of baselines contributing to the sensitivity of a power spectrum estimate. In \citet{ali_et_al2015}, this number was the total number of $30$ m East/West baselines used in the analysis. However, using the total number of baselines neglects the fact that we divid baselines into $5$ groups before cross-multiplying data. Our revised estimate for the parameter is $N_{bls} = \frac{N_{bls}}{N_{gps}}\sqrt{N_{gps}^{2}-N_{gps}}$, where $N_{gps} = 5$. This expression arises because $N_{bls}$ must be scaled by the total number of cross-multiplications going into the power spectrum estimate. For $5$ groups, there are $25$ total cross-products, minus $5$ cross-products since we avoid multiplying the same group with itself. For our PAPER-64 analysis with only one baseline separation type, this becomes $N_{bls} \sim 46$ baselines. 
\item $N_{pols}$: The number of polarizations averaged together. For the case of Stokes I, $N_{pols}=2$.
\end{itemize}
